---
title: "Guinea worm mitochondrial DNA genomics - calculating nucleotide similarity"
output: html_notebook
---

## Load libraries, scripts, and data

```{r setup}
# load libraries
for(p in c('data.table', 'dplyr', 'tidyr', 'protr', 'geosphere', "ggplot2")){
  if(!p %in% installed.packages()[,1]){
    install.packages(p)
    library(p, character.only = TRUE)
  }
  library(p, character.only = TRUE)
}

# set global options
options(datatable.fread.datatable=FALSE)
options(stringsAsFactors = FALSE)

# set directories
code_dir <- "/mnt/md0/guinea_worm/mt_sanger"
project_dir <- "/home/jribado/Dropbox (IDM)/Data, Dynamics, and Analytics Folder/Projects/Guinea Worm Genetics in Chad/gw_bwaAlign/2020_newSeqs"

# load functions
source(paste(code_dir, "plotting_themeJ.R", sep="/"))
source(paste(code_dir, "genSim_functions.R", sep="/"))
```

```{r load_data}
# load barcodes 
barcode_numbers <- read.delim(paste(project_dir, "202004_bc2bciden.txt", sep="/"), sep="\t")
barcode_vars <- read.delim(paste(project_dir, "202003_geneBarcodes.txt", sep="/"), sep="\t")
# create a sequenced name data frame to merge with national line lists 
seq_worms <- cbind.data.frame(
  worm = barcode_vars$worm,
  year = as.integer(sapply(barcode_vars$worm, function(i){gsub(".*2", "2", unlist(strsplit(i, "_")))[1]})),
  worm_number = sapply(barcode_vars$worm, function(i){unlist(strsplit(i, "_"))[4]}),
  host_number = sapply(barcode_vars$worm, function(i){gsub("\\..*", "", unlist(strsplit(i, "_"))[4])})
)

# load national line and case merged data
worm_meta <- data.table::fread('/home/jribado/Dropbox (IDM)/Data, Dynamics, and Analytics Folder/Projects/Guinea Worm Genetics in Chad/sample_identifiers/202004_natLineCasesDuplicates.txt', sep="\t") 
# merge
# seq_metaMerge <- dplyr::left_join(seq_worms, worm_meta)
# write.table(seq_metaMerge, paste(project_dir, "202004_wormMetadata.txt", sep="/"), sep="\t", quote = F, row.names = F)
# ^ added some manual editing to the file 
seq_metaMerge <- read.delim(paste(project_dir, "202004_wormMetadata.txt", sep="/"), sep="\t")
```

## Calculate nucleotide similarity (Pi metric)

05/2020: This is incorrect because it uses protein matrices to determine subsitution penalties - which are not the same for each nucelotide. Scratch this, and just got it manually for each pair. 

```{r sing_inc, eval=F}
bc_list <- df2barcodeList(dplyr::select(barcode_vars, worm, full_barcode))
bc_mat <- simMat(bc_list, 16 , names(bc_list))
bc_df  <- simMat2df(bc_mat)
write.table(bc_df, paste(project_dir, "202004_genePairwise.txt", sep="/"), sep="\t", quote = F, row.names = F)
```

```{r gen_sim, eval=F}
  string_diff <- function(a, b){
    if(nchar(a)!=nchar(b)) stop("Lengths of input strings differ. Please check your input.")
    seq.a <- unlist(strsplit(a,split=""))
    seq.b <- unlist(strsplit(b,split=""))
    diff.d <- rbind(seq.a,seq.b)
    only.diff <-diff.d[,diff.d[1,]!=diff.d[2,]]
    pos <- which(diff.d[1,]!=diff.d[2,])
    return(1-(length(pos)/nchar(a)))
  }

barcodes <- unique(barcode_vars$full_barcode)
barcode_groups <- data.frame(gtools::permutations(n=length(barcodes),r=2,v=barcodes, repeats.allowed=TRUE)) 
similarity <- unlist(lapply(seq(1, nrow(barcode_groups)), function(i){
  string_diff(barcode_groups$X1[i], barcode_groups$X2[i])
}))
barcode_sim <- cbind.data.frame(barcode_groups, similarity) 
names(barcode_sim) <- c("full_barcode.x", "full_barcode.y", "value")
write.table(barcode_sim, paste(project_dir, "20200515_bcPairwiseManual.txt", sep="/"), sep="\t", quote = F, row.names = F)
```

## Combine similairty scores with different worm pairs 

```{r merge_metadata}
# load worms we believe have incorrect coordinates
gps_out <- readRDS(paste(project_dir, "GPS_outlierHosts.Rds", sep="/")) 

# create a master list with metadata
worm_metaMerge <- seq_metaMerge[,1:8] %>% unique() %>% 
  dplyr::left_join(., dplyr::select(barcode_vars, worm, full_barcode)) %>%
  dplyr::left_join(., barcode_numbers) %>%
  dplyr::mutate(incorrect_gps = ifelse(host_number %in% gps_out, paste(latitude, longitude, sep=","), NA),
                latitude = ifelse(host_number %in% gps_out, NA, latitude),
                longitude = ifelse(host_number %in% gps_out, NA, longitude))
# write.table(worm_metaMerge, paste(project_dir, "20200515_wormMeta.txt", sep="/"), sep="\t", quote = F, row.names = F)

# create  
worm_pairs <- data.frame(t(combn(worm_metaMerge$worm,2))) %>% dplyr::rename(worm1 = X1, worm2 = X2) %>%
  dplyr::left_join(., worm_metaMerge, by=c("worm1" = "worm")) %>%
  dplyr::left_join(., worm_metaMerge, by=c("worm2" = "worm")) 
dist_m <- sapply(1:nrow(worm_pairs), function(x){
  geosphere::distm(as.numeric(worm_pairs[x,8:9]), 
                   as.numeric(worm_pairs[x,18:19]), fun=distHaversine)}) 
gw_genSim <- cbind.data.frame(worm_pairs, dist_m, stringsAsFactors=F) %>%
  dplyr::left_join(., barcode_sim) %>%
  dplyr::mutate(bc_match = ifelse(full_barcode.x == full_barcode.y, "True", "False"),
                yr_match = ifelse(year.x == year.y, "True", "False"),
                host_match = ifelse(host_number.x == host_number.y, "True", "False"),
                species_match = ifelse((grepl("HUM", worm1) & grepl("HUM", worm2)) | (grepl("DOG", worm1) & grepl("DOG", worm2)), "True", "False"))
# write.table(gw_genSim, paste(project_dir, "20200515_bcPairwiseManualMeta.txt", sep="/"), sep="\t", quote = F, row.names = F)
```

```{r merge_metadata_old, eval=F}
bc_df <- data.table::fread(paste(project_dir, "202004_genePairwise.txt", sep="/"), sep="\t")
sim_merge <- pairwise_comp(bc_df, seq_metaMerge, barcode_vars) 
dist_m <- sapply(1:nrow(sim_merge), function(x){
  geosphere::distm(as.numeric(sim_merge[x,9:10]), 
                   as.numeric(sim_merge[x,27:28]), fun=distHaversine)}) 
gw_genSim <- cbind.data.frame(sim_merge, dist_m, stringsAsFactors=F)
# write.table(gw_genSim, paste(project_dir, "202004_genePairwiseMetadata.txt", sep="/"), sep="\t", quote = F, row.names = F)
```

## Phylogenetic trees

```{r phy_trees}
bc_mat <- do.call(rbind, strsplit(barcode_vars$full_barcode, ""))
rownames(bc_mat) <- barcode_vars$worm
# create phylogenetic object
gw_phydat <- phangorn::phyDat(bc_mat)
# calculate distances
gw_dist   <- phangorn::dist.ml(gw_phydat)
# create trees
gw_UPGMA  <- phangorn::upgma(gw_dist)
print(paste("UPGMA parsimony:", phangorn::parsimony(gw_UPGMA, gw_phydat)))
gw_NJ     <- phangorn::NJ(gw_dist)
print(paste("Nearest neighbor joining parsimony:", phangorn::parsimony(gw_NJ, gw_phydat)))
```

From maximum parsimony calculations, UPGMA is a better tree. As the simplest method to construct tree, it's assumptions that all lineages have the same evolutonary pressure holds since it's the same organism. 

```{r plot_phylogenies}
ggtree::ggtree(gw_UPGMA, branch.length = "none")
```

This tree is a little budy, with large clades per barcode. To simplify this, I will take one representative for each barcode per species and plot the number of cases per year as a dotplot next to the tree. 

```{r species_barcode}
barcode_counts <- read.delim('/home/jribado/Dropbox (IDM)/Data, Dynamics, and Analytics Folder/Projects/Guinea Worm Genetics in Chad/gw_bwaAlign/2020_newSeqs/202004_bc2bciden.txt')

bc_meta <- dplyr::mutate(barcode_vars, 
  host = ifelse(grepl("HUM", worm), "Human", "Dog"),
  year = stringr::str_extract(sub('\\_.*', '', worm), "\\-*\\d+\\.*\\d*")) %>%
  dplyr::left_join(., barcode_counts)
bc_species <- dplyr::group_by(bc_meta, full_barcode, host, year, barcode_number) %>% 
  dplyr::summarise(count=n()+5) %>%
  tidyr::pivot_wider(names_from = year, values_from = count) %>%
  replace(., is.na(.), 0)
# create new barcode matrix to build phylogenetic tree
bc_spec_mat <- do.call(rbind, strsplit(bc_species$full_barcode, ""))
rownames(bc_spec_mat) <- paste(bc_species$host, bc_species$barcode_number, sep="_")

# create phylogenetic object
spec_phydat <- phangorn::phyDat(bc_spec_mat)
# calculate distances
spec_dist   <- phangorn::dist.ml(spec_phydat)
# create tree
spec_UPGMA  <- phangorn::upgma(spec_dist)
```

```{r plot_phylogenies_simp}
side_map <- as.matrix(bc_species[,c(9, 8, 7, 5, 6, 4)])
side_map[side_map > 1 ] <- 1
row.names(side_map) <- paste(bc_species$host, bc_species$barcode_number, sep="_")
# identify colors
sort(unique(as.vector(side_map)))
# phytools::dotTree(spec_UPGMA, side_map, labels=TRUE, length=19, fsize=0.7, ftype="i")
png(paste(project_dir, "plots", "20200515_phylo_quantCase.png", sep="/"), width = 600, height = 800)
phytools::phylo.heatmap(spec_UPGMA, side_map, length=100, 
                         length=2, fsize=0.7, ftype="i", colors=c("grey99", "navyblue"))
                         # colors=c("grey99", 
                         #         '#fef0d9','#fdd49e','#fdbb84','#fc8d59','#ef6548','#d7301f','#990000',
                         #         '#edf8e9','#c7e9c0','#a1d99b','#74c476','#31a354','#006d2c',
                         #         '#c6dbef','#9ecae1','#6baed6','#3182bd','#08519c'))
dev.off()
```

## Compare human and dog similarity

```{r species_sim}
gw_comp <- data.table::fread(paste(project_dir, "20200515_bcPairwiseManualMeta.txt", sep="/"), sep="\t") %>%
  dplyr::filter(species_match == "FALSE") %>%
  dplyr::mutate(bc_iden.x = ifelse(grepl("HUM", worm1), paste("Human", barcode_number.x), paste("Dog", barcode_number.x)),
                bc_iden.y = ifelse(grepl("HUM", worm2), paste("Human", barcode_number.y), paste("Dog", barcode_number.y)))
bc_compIden <- dplyr::select(gw_comp, bc_iden.x, bc_iden.y, value) %>% unique() %>%
  dplyr::mutate(bc_iden.x = factor(bc_iden.x, levels = rev(gtools::mixedsort(unique(bc_iden.x)))),
                bc_iden.y = factor(bc_iden.y, levels = gtools::mixedsort(unique(bc_iden.y))))
```

```{r species_sim_plot}
coord_ratio <- length(unique(bc_compIden$bc_iden.x))/length(unique(bc_compIden$bc_iden.y)) 
bc_iden <- bc_compIden %>% ggplot(aes(x=bc_iden.y, y=bc_iden.x, fill=value)) + 
  geom_tile() +
  scale_fill_viridis_c(name="Genetic\nsimilarity") + 
  scale_x_discrete(position = "top") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        axis.title = element_blank()) +
  coord_flip()
  #coord_fixed(ratio = coord_ratio)
ggsave("20200515_species_pairwiseIden.png", plot = bc_iden, path = paste(project_dir, "plots", sep="/"), width = 6, height = 6, units = c("in"), dpi = 300) 
```
